---
title: "Lab 3: Data Storytelling and Visualization in R"
author: ""
format: html
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

# Overview

This lab explores data storytelling using R through: - Exploratory data
analysis (EDA) - Transformations and skewness - Regression
visualization - Time series anomalies - Composite plots - Interactive
visualization (Shiny) - A capstone narrative

------------------------------------------------------------------------

# Setup

```{r}
# Install if needed
# install.packages(c("tidyverse", "moderndive", "fivethirtyeight",
#                    "patchwork", "scales", "shiny"))

library(tidyverse)
library(moderndive)
library(fivethirtyeight)
library(patchwork)
library(scales)
```

------------------------------------------------------------------------

# Exercise 1: Exploratory Data Analysis & Skewness

## Raw Distribution

```{r}
data("house_prices")

p1_raw <- ggplot(house_prices, aes(x = price)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  scale_x_continuous(labels = label_dollar(scale = 1e-6, suffix = "M")) +
  labs(
    title = "Distribution of House Prices (Raw)",
    subtitle = "Right-skewed distribution",
    x = "Price (USD)",
    y = "Count"
  ) +
  theme_bw()

p1_raw
```

## Log Transformation

```{r}
house_prices <- house_prices %>%
  mutate(
    log10_price = log10(price),
    log10_size  = log10(sqft_living)
  )

p1_log <- ggplot(house_prices, aes(x = log10_price)) +
  geom_histogram(bins = 50, fill = "darkorange", color = "white") +
  labs(
    title = "Distribution of House Prices (log10 scale)",
    subtitle = "More symmetric distribution",
    x = "log10(Price)",
    y = "Count"
  ) +
  theme_bw()

p1_log
```

### Interpretation

The raw distribution is heavily right-skewed due to high-value
outliers.\
The log transformation reveals a more symmetric, near-normal
distribution, improving interpretability.

------------------------------------------------------------------------

# Exercise 2: Interaction vs Parallel Slopes

```{r}
house_prices <- house_prices %>%
  mutate(condition = factor(condition, levels = 1:5,
                            labels = c("Poor","Fair","Average","Good","Excellent")))

house_sub <- house_prices %>%
  filter(condition %in% c("Average", "Good", "Excellent"))
```

## Interaction Model

```{r}
p2_interaction <- ggplot(house_sub,
                         aes(x = log10_size, y = log10_price, color = condition)) +
  geom_point(alpha = 0.15) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Interaction Model") +
  theme_bw()

p2_interaction
```

## Parallel Slopes Model

```{r}
p2_parallel <- ggplot(house_sub,
                      aes(x = log10_size, y = log10_price, color = condition)) +
  geom_point(alpha = 0.15) +
  moderndive::geom_parallel_slopes(se = FALSE) +
  labs(title = "Parallel Slopes Model") +
  theme_bw()

p2_parallel
```

### Interpretation

Slopes are similar across conditions, supporting a simpler parallel
slopes interpretation: - Size affects price consistently - Condition
shifts baseline price

------------------------------------------------------------------------

# Exercise 3: Time Series Anomaly

```{r}
data("US_births_1994_2003")

births_1999 <- US_births_1994_2003 %>%
  filter(year == 1999) %>%
  mutate(date = as.Date(paste(year, month, date_of_month, sep = "-")))
```

```{r}
ggplot(births_1999, aes(x = date, y = births)) +
  geom_line(color = "steelblue") +
  labs(title = "Daily Births in 1999")
```

## Peak Detection

```{r}
births_1999 %>%
  arrange(desc(births)) %>%
  head()
```

### Interpretation

The largest spike occurs on **September 9, 1999 (9/9/99)**, likely due
to cultural preferences influencing scheduled births.

------------------------------------------------------------------------

# Exercise 4: Composite Visualization

## Plot A

```{r}
p4a <- ggplot(house_sub,
              aes(x = condition, y = log10_price, fill = condition)) +
  geom_boxplot() +
  theme_bw()
```

## Plot B

```{r}
p4b <- ggplot(house_sub,
              aes(x = log10_size, y = log10_price,
                  color = condition, shape = condition)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw()
```

## Combined

```{r}
p4a | p4b
```

------------------------------------------------------------------------

# Exercise 5: Shiny App

> Run separately as `app.R`

```{r}
# shinyApp(ui, server)
```

(App code omitted in rendering but included in source file)

------------------------------------------------------------------------

# Capstone: What Drives House Prices?

## Step 1: Distribution

```{r}
cap1 <- p1_raw / p1_log
cap1
```

## Step 2: Condition Premium

```{r}
median_by_cond <- house_prices %>%
  group_by(condition) %>%
  summarise(median_price = median(price), .groups = "drop")

cap2 <- ggplot(median_by_cond,
               aes(x = condition, y = median_price, fill = condition)) +
  geom_col() +
  scale_y_continuous(labels = label_dollar()) +
  theme_bw()

cap2
```

## Step 3: Size + Condition

```{r}
cap3 <- ggplot(house_sub,
               aes(x = log10_size, y = log10_price, color = condition)) +
  geom_point(alpha = 0.2) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_bw()

cap3
```

## Final Narrative

```{r}
cap1 | (cap2 / cap3)
```

------------------------------------------------------------------------

# Conclusions

- House prices are **right-skewed**, requiring transformation.
- **Condition strongly affects price levels**.
- **Size increases price consistently across conditions**.

## Call to Action

- Sellers should invest in condition improvements.
- Buyers can find value in larger homes needing upgrades.

------------------------------------------------------------------------

# Appendix: Shiny App Code

```{r}
library(shiny)

house_shiny <- house_prices %>%
  filter(condition %in% c("Average", "Good", "Excellent"))

ui <- fluidPage(
  titlePanel("House Prices Explorer"),
  sidebarLayout(
    sidebarPanel(
      selectInput("condition_select", "Condition:",
                  choices = c("Average", "Good", "Excellent"))
    ),
    mainPanel(plotOutput("scatter_plot"))
  )
)

server <- function(input, output) {
  filtered_data <- reactive({
    house_shiny %>% filter(condition == input$condition_select)
  })

  output$scatter_plot <- renderPlot({
    ggplot(filtered_data(),
           aes(x = log10_size, y = log10_price)) +
      geom_point(alpha = 0.3) +
      geom_smooth(method = "lm") +
      theme_bw()
  })
}

# shinyApp(ui, server)
```

::: callout-important
## Work task

For the final part of the lab, you must produce a coherent data story
that includes:

1.  **At least three different plot types**.
2.  **Proper annotation** (titles, axis labels with units, and
    captions).
3.  **A call to action** or a concluding summary of what the visuals
    reveal.
:::
